---
title: Extracting Business Cycles from Time Series Data
author: Nikhil Damodaran
date: '2020-08-10'
math: yes
description: a blog which discusses extraction of business cycles from time series
  data
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<p class="red"> __Getting Started__ </p>

We start with the basics of R, exploring the memory and clearing out backlog from previous work.

```{r, eval = FALSE}
ls() # checking R's memory...
list.files() # checking files in my working directory...
# rm(list = ls()) # clearing all previously stored items...
# import data from your excel sheet
install.packages("readxl") # for importing from excel
```

Then we import the data from an excel sheet.

```{r}
library(readxl)
gdp <- read_excel("/Users/nikhildamodaran/Dropbox/Nikhil/teaching/Macroeconomics-II/Fall-2020/Teaching Resources/Exercise 1/GDP FRED USA.xlsx")

# features of the data
class(gdp) 
str(gdp) # structure of the data
head(gdp) # first 5 rows...
```

Tell me the features of the variable

```{r}
sapply(gdp, class)
```

Note that first variable is POSIXct - a time series object.

```{r}
ts.plot(gdp$gdp.usd)
```

<p class="red"> __Converting Variables into Time Series Objects__ </p>

Problem: GDP variable is not time series but only a bunch of numbers

```{r , eval = TRUE}
# converting the vector into time series with a starting and ending time
gdp$gdp.usd <- ts(data=gdp$gdp.usd, frequency = 4,start=c(1947,1), end=c(2020,2)) 

# checking the class of variables in data
sapply(gdp, class)

class(gdp$gdp.usd)
```

Now plotting this in R again:

```{r , eval = TRUE}
ts.plot(gdp$gdp.usd)
# tell me the difference between the first and the second plot
```

Problems remain: axis is not marked properly for reading.

<p class="red"> __Removing Seasonality__ </p>

We need to undergo seasonal transformation only if the data is not seasonalized. This is not a problem for annual data, but for lower frequencies we need to remove the effects of seasonal fluctuations. 

```{r , eval = FALSE}
# Lets extract business cycles from US data
# Step 1: Remove seasonal effects:
install.packages("seasonal") # for seasonal 
```

Remove seasonal effects such as diwali:

```{r , eval = TRUE}
library(seasonal)
deseasonal.gdp <- seas(gdp$gdp.usd)
```

Check what are the results of seasonal adjustment and understand what you have done:

```{r , eval = TRUE}
summary(deseasonal.gdp) # you've done a regression
attributes(deseasonal.gdp) # get the data 
library(ggplot2)
gdp$deseasonal <- deseasonal.gdp$data[,3] # store this data in your dataset
ggplot(gdp, aes (y=gdp$gdp.usd, x = gdp$time)) + geom_line(colour=alpha('blue', 0.85)) + geom_line(aes(y=gdp$deseasonal, x = gdp$time), colour=alpha('red', 0.85) )  + theme_minimal() + xlab("Original (blue) and De-Seasonalized (red)") + ggtitle("Removing Seasonal Fluctuations from US GDP") + ylab("Bn. USD")
```

<p class="red"> __Isolating Business Cycles__ </p>

Installing packages required for filtering

```{r , eval = FALSE}
# Step 2: Lets get the business cycles
install.packages("mFilter")
```

Getting the business cycles:

```{r , eval = TRUE}
library(mFilter)
usa.business.cycles <- hpfilter(log(gdp$deseasonal), freq = 1600)
attributes(usa.business.cycles)
ts.plot(usa.business.cycles$cycle, gpars = list(ylab ="US GDP Cycles"))
```

Obtaining the summary statistics for these business cycles

```{r}
summary(usa.business.cycles$cycle)
sd(usa.business.cycles$cycle) # standard deviation of GDP cycles
acf(usa.business.cycles$cycle) # auto-correlation function
acf(usa.business.cycles$cycle, plot = FALSE)
```
